Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-04-15 src/tex/ms.tex
added loss figure and other updates

In this paper, we investigate the possibility of further improving the accuracy
of IMRPhenomD by jointly optimizing all the fitting coefficients given NR
waveforms, and what constraints one may face when trying to further improve. We
find that simply by applying gradient descent algorithm, one can obtain a better
set of waveform coefficients, thus improving the accuracy of the model.
Furthermore, by comparing the accuracy of optimized and original waveforms, we
find that model-generated waveforms are very sensitive to their intrinsic
parameters. Specifically, IMRPhenomD favors certain parts of the parameter
space. This means IMRPhenomD introduces systematic bias to other GW analysis
tasks. This showcases the flaws of the ansatz and allows us to have a deeper
understanding of Phenom models.  
% \kw{Briefly describe the high level finding in this paper}

@@

% Need to change the focus to flat PSD instead of ALIGO PSD
Other than NR waveforms, one need to choose a relevant noise PSD for 
mismatch. We have opted to use a flat PSD for this purpose, as it provides results that are independent of the detector sensitivity and mass scale. The use of a flat PSD ensures that the improvement in accuracy is due solely to the difference in high-dimensional fitting and piece-wise fitting. Furthermore, we are interested in examining the effect of introducing a detector PSD on the optimization process. For this, we have chosen the zero-detuned high-power (\zdethp) noise PSD \citep{aasi2015advanced}. Since the total mass of the system scales with the frequency of the waveform, we must choose a corresponding mass scale to match the frequency range of our noise PSD. Therefore, we have selected a mass scale of $M=50M_{\odot}$ for all waveforms, as this is a commonly observed mass scale in LIGO observations. We will then examine the differences in optimization.

@@

\begin{figure}
	\script{loss.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/loss.pdf}
%	\caption{Parameter space with mass ratio against normalized reduced spin. Orange: Training waveforms; Blue: Testing waveforms}
	\label{fig:loss}
\end{figure}

@@

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, whereas spinning waveforms can either improve or
worsen in terms of mismatch. \kw{What does this means?}

Furthermore, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e.
$\chi_1\approx\chi_2$, show good mismatch improvements. We expect to see this
feature since the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. One
interesting feature is how the second and fourth quadrants respond to
optimization. \kw{How did you count quadrant?} In the second quadrant ($\chi_1<0$ and $\chi_2>0$), waveforms
generally improve with along optimization. However, mismatches in the fourth
quadrant ($\chi_1>0$ and $\chi_2<0$) do not improve after optimization. Most
waveforms even turned worse after optimization. These waveforms correspond to
the waveforms in the high mismatch tail in Fig.~\ref{fig:q148}. \kw{Give justification.}

@@

\begin{table}[t]
	\centering
	\begin{tabularx}{0.8\columnwidth}{@{\extracolsep{\fill}}lrrr}
		\toprule\midrule Code         & $q$ & $\chi_1$ & $\chi_2$ \\
		\midrule\midrule SXS:BBH:0172 & 1.0 & 0.98     & 0.98     \\
		SXS:BBH:0152 & 1.0 & 0.60     & 0.60     \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1417 & 4.0 & 0.40     & 0.50     \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1426 & 8.0 & 0.48     & 0.75     \\
		SXS:BBH:0167 & 8.0 & 0.00     & 0.00     \\ \midrule SXS:BBH:0370 & 1.0
		& -0.20    & 0.40     \\
		SXS:BBH:2092 & 1.0 & -0.50    & 0.50     \\
		SXS:BBH:0330 & 1.0 & -0.80    & 0.80     \\
		SXS:BBH:2116 & 2.0 & -0.30    & 0.30     \\
		SXS:BBH:2111 & 2.0 & -0.60    & 0.60     \\
		SXS:BBH:0335 & 2.0 & -0.80    & 0.80     \\
		SXS:BBH:0263 & 3.0 & -0.60    & 0.60     \\
		SXS:BBH:2133 & 3.0 & -0.73    & 0.85     \\
		SXS:BBH:0263 & 4.0 & -0.80    & 0.80     \\ \midrule SXS:BBH:0156 & 1.0
		& -0.95    & -0.95    \\
		SXS:BBH:0151 & 1.0 & -0.60    & -0.60    \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1418 & 4.0 & -0.40    & -0.50    \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1419 & 8.0 & -0.80    & -0.80    \\
		SXS:BBH:0063 & 8.0 &  0.00    &  0.00    \\ \midrule SXS:BBH:0304 & 1.0
		& 0.50     & -0.50    \\
		SXS:BBH:0327 & 1.0 & 0.80     & -0.80    \\
		SXS:BBH:2123 & 2.0 & 0.30     & -0.30    \\
		SXS:BBH:2128 & 2.0 & 0.60     & -0.60    \\
		SXS:BBH:2132 & 2.0 & 0.87     & -0.85    \\
		SXS:BBH:2153 & 3.0 & 0.30     & -0.30    \\
		SXS:BBH:0045 & 3.0 & 0.50     & -0.50    \\
		SXS:BBH:0292 & 3.0 & 0.73     & -0.85    \\ \midrule\bottomrule
	\end{tabularx}
	\caption{List of waveforms used in recalibrating coefficients in 4 quadrants. From top to down are the first, second, third and fourth quadrants. Note that for the first and third quadrants, waveforms are chosen to have equal or similar spins, while the training waveforms for the second and fourth quadrants are chosen to have opposite spins.}
	\label{tab:quadrants}
\end{table}

We divided the parameter space into four quadrants to analyze the effect of the
recalibration procedure on each region separately (Fig.\ref{fig:ps_q148}). The
training waveforms used for fitting are listed in Table \ref{tab:quadrants}. The
second and fourth quadrants have limited data for $q>4$, hence the results are
only valid up to $q\leq4$. From Fig.\ref{fig:ps_q148_quadrant}, we observe
improvements in waveform mismatch for all waveforms, except those in the fourth
quadrant. For equal-spin waveforms lying in the proximity of the diagonal axis,
the mismatch improves significantly due to most training waveforms being
equal-spin. In the second quadrant, the improvement is noticeable, except for a
few defects caused by some testing waveforms having $q>4$. The optimized
waveforms in the fourth quadrant have a higher mismatch than the original ones,
indicating poor fitting. Further optimization in a smaller subspace does not
improve the results.

\begin{figure}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches after optimizing in separate quadrants.
	We use a constant noise spectrum to calculate mismatch and
	$\mathcal{L}_{\mathrm{mean}}$ for the loss function. Generally, most
	waveforms with mismatch $>10^{-2}$ lies in the fourth quadrant. }
	\label{fig:all_quadrants}
\end{figure}

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{\kw{The color scheme makes the plot a bit hard to read.}Parameter
	space of testing waveforms. Each quadrant is fitted independently. Colorbar
	represents log difference of mismatches before and after optimization.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

The reduced spin approximation is a major contributor to the inaccuracies
observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
parameter, $\chi_{\mathrm{PN}}$, as described in Sec. \ref{sec:method}. The
parameterization of BBH mergers using a single spin parameter results in a
degeneracy within the parameter space. Specifically, black hole events with
different spins may generate the same waveform due to identical values of
$\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
unequal spin events. This degeneracy produces straight lines in the parameter
space with negative slopes that depend on the mass ratio, as shown in
Fig.\ref{fig:ps_q148}. Notably, the ansatz performs better in the top left
region than the bottom right along a degeneracy line. In an attempt to address
this issue, we partitioned the parameter space into four quadrants, as described
in Sec. \ref{sec:result}. However, even with separate optimizations for each
quadrant, Fig.\ref{fig:ps_q148_quadrant} indicates that the fourth quadrant
exhibits similar mismatches as before, while the second quadrant's performance
has improved. This observation suggests that the ansatz is specific to certain
regions of the parameter space, with a preference for BBH events featuring
$\chi_1<0$ and $\chi_2>0$.

The division of the parameter space into four quadrants was a simple approach
taken for practical reasons. A more systematic approach would involve the use of
level set estimation algorithms to identify regions of interest within the
parameter space. Such an algorithm can reveal additional degeneracies or issues
that may exist within the ansatz. One possible strategy is to recalibrate
individual regions of interest to achieve better results. An alternative
approach is to select regions based on degeneracy lines. However, due to the
limited number of NR waveforms available, we were unable to implement this
approach. With more NR waveforms available in the future that cover the entire
parameter space, we can perform optimization with fewer restrictions and select
regions more systematically.

@@

Furthermore, we investigated how the intrinsic parameters affect the
improvement. Fig.~\ref{fig:ps_q148} shows that the optimization procedure has a
certain preference for waveforms lying in the second quadrant while the fourth
quadrant cannot be improved. To further test this result, we recalibrate
waveforms in separate regions in parameter space. As shown in
Fig.~\ref{fig:ps_q148_quadrant}, this recalibration process gives further
improvement to the second quadrant while the fourth quadrant shows similar
result. This indicates that the model's ansatz does not fit waveforms in the
fourth quadrant. This phenomenon is due to the reduced spin approximation used
in parameterizing the ansatz, where degeneracies between $\chi_1$ and $\chi_2$
are introduced.

While we naively separate the optimization process into 4 quadrants, one can
perform systematic region-selection. In principle, we can apply this general
method to other newer and more accurate models such as IMRPhenomX or IMRPhenomP
models. Then, we can perform all the above analyses to understand how to
construct better GW Phenom models in the future.

>>>>

% Put more focus on the joint optimization instead of having a set of more accurate 
% coefficients, since the method is the main point
In this paper, we investigate the possibility of further improving the accuracy
of IMRPhenomD by jointly optimizing all the fitting coefficients given NR
waveforms, and what constraints one may face when trying to further improve. We
find that simply by applying gradient descent algorithm, one can obtain a better
set of waveform coefficients, thus improving the accuracy of the model.
Furthermore, by comparing the accuracy of optimized and original waveforms, we
find that model-generated waveforms are very sensitive to their intrinsic
parameters. Specifically, IMRPhenomD favors certain parts of the parameter
space. This means IMRPhenomD introduces systematic bias to other GW analysis
tasks. This showcases the flaws of the ansatz and allows us to have a deeper
understanding of Phenom models.

@@

% Need to change the focus to flat PSD instead of ALIGO PSD
Other than NR waveforms, one need to choose a relevant noise PSD for 
mismatch. We have opted to use a flat PSD for this purpose, as it provides results that are independent of the detector sensitivity and mass scale. The use of a flat PSD ensures that the improvement in accuracy is due solely to the difference in high-dimensional fitting and piece-wise fitting. Furthermore, we are interested in examining the effect of introducing a detector PSD on the optimization process. For this, we have chosen the zero-detuned high-power (\zdethp) noise PSD \citep{aasi2015advanced}. Since the total mass of the system scales with the frequency of the waveform, we must choose a corresponding mass scale to match the frequency range of our noise PSD. We selected a mass scale of $M=50M_{\odot}$ for all waveforms, as this is a commonly observed mass scale in LIGO observations. We will then examine the differences in optimization.

@@

\begin{figure}
	\script{loss.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/loss.pdf}
	\caption{The value of loss function against number of iterations. The code is terminated after 30000 steps }
	\label{fig:loss}
\end{figure}

@@

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, whereas spinning waveforms can either improve or
worsen in terms of mismatch base on their intrinsic parameters. 
%\kw{What does this means?}

Furthermore, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e.
$\chi_1\approx\chi_2$, show good mismatch improvements. We expect to see this
feature since the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. One
interesting feature is how the top-left ($\chi_1<0$ and $\chi_2>0$) and bottom-right ($\chi_1>0$ and $\chi_2<0$) regions respond to
optimization. In the top-left region, waveforms
generally improve with along optimization. However, mismatches in the bottom-right region do not improve after optimization. Most
waveforms even turned worse after optimization. These waveforms correspond to
the waveforms in the high mismatch tail in Fig.~\ref{fig:q148}. \kw{Give justification.}

@@

\begin{table}[t]
	\centering
	\begin{tabularx}{0.8\columnwidth}{@{\extracolsep{\fill}}lrrr}
		\toprule\midrule Code         & $q$ & $\chi_1$ & $\chi_2$ \\
		\midrule\midrule SXS:BBH:0172 & 1.0 & 0.98     & 0.98     \\
		SXS:BBH:0152 & 1.0 & 0.60     & 0.60     \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1417 & 4.0 & 0.40     & 0.50     \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1426 & 8.0 & 0.48     & 0.75     \\
		SXS:BBH:0167 & 8.0 & 0.00     & 0.00     \\ \midrule SXS:BBH:0370 & 1.0
		& -0.20    & 0.40     \\
		SXS:BBH:2092 & 1.0 & -0.50    & 0.50     \\
		SXS:BBH:0330 & 1.0 & -0.80    & 0.80     \\
		SXS:BBH:2116 & 2.0 & -0.30    & 0.30     \\
		SXS:BBH:2111 & 2.0 & -0.60    & 0.60     \\
		SXS:BBH:0335 & 2.0 & -0.80    & 0.80     \\
		SXS:BBH:0263 & 3.0 & -0.60    & 0.60     \\
		SXS:BBH:2133 & 3.0 & -0.73    & 0.85     \\
		SXS:BBH:0263 & 4.0 & -0.80    & 0.80     \\ \midrule SXS:BBH:0156 & 1.0
		& -0.95    & -0.95    \\
		SXS:BBH:0151 & 1.0 & -0.60    & -0.60    \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1418 & 4.0 & -0.40    & -0.50    \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1419 & 8.0 & -0.80    & -0.80    \\
		SXS:BBH:0063 & 8.0 &  0.00    &  0.00    \\ \midrule SXS:BBH:0304 & 1.0
		& 0.50     & -0.50    \\
		SXS:BBH:0327 & 1.0 & 0.80     & -0.80    \\
		SXS:BBH:2123 & 2.0 & 0.30     & -0.30    \\
		SXS:BBH:2128 & 2.0 & 0.60     & -0.60    \\
		SXS:BBH:2132 & 2.0 & 0.87     & -0.85    \\
		SXS:BBH:2153 & 3.0 & 0.30     & -0.30    \\
		SXS:BBH:0045 & 3.0 & 0.50     & -0.50    \\
		SXS:BBH:0292 & 3.0 & 0.73     & -0.85    \\ \midrule\bottomrule
	\end{tabularx}
	\caption{List of waveforms used in recalibrating coefficients in 4 regions. From top to down are the top-right ($\chi_1,\chi_2>0$), top-left ($\chi_1<0<\chi_2$), bottom-left ($\chi_1,\chi_2<0$) and bottom-right ($\chi_1>0>\chi_2$) regions. Note that for the top-right and bottom-left regions, waveforms are chosen to have equal or similar spins, while the training waveforms for the other two regions are chosen to have opposite spins.}
	\label{tab:quadrants}
\end{table}

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately (Fig.\ref{fig:ps_q148}). The
training waveforms used for fitting are listed in Table \ref{tab:quadrants}. The
top-left and bottom-right regions have limited data for $q>4$, hence the results are
only valid up to $q\leq4$. From Fig.\ref{fig:ps_q148_quadrant}, we observe
improvements in waveform mismatch for all waveforms, except those in the bottom-right region. For equal-spin waveforms lying in the proximity of the diagonal axis,
the mismatch improves significantly due to most training waveforms being
equal-spin. In the top-left region, the improvement is noticeable, except for a
few defects caused by some testing waveforms having $q>4$. The optimized
waveforms in the bottom-right region have a higher mismatch than the original ones,
indicating poor fitting. Further optimization in a smaller subspace does not
improve the results.

\begin{figure}[t]
	\script{all_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/all_quadrants.pdf}
	\caption{Distributions of mismatches after optimizing in separate regions.
	We use a constant noise spectrum to calculate mismatch and
	$\mathcal{L}_{\mathrm{mean}}$ for the loss function. Generally, most
	waveforms with mismatch $>10^{-2}$ lies in the bottom-right region. }
	\label{fig:all_quadrants}
\end{figure}

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{\kw{The color scheme makes the plot a bit hard to read.}Parameter
	space of testing waveforms. Each region is fitted independently. Colorbar
	represents log difference of mismatches before and after optimization.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

The reduced spin approximation is a major contributor to the inaccuracies
observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
parameter, $\chi_{\mathrm{PN}}$, as described in Sec. \ref{sec:method}. The
parameterization of BBH mergers using a single spin parameter results in a
degeneracy within the parameter space. Specifically, black hole events with
different spins may generate the same waveform due to identical values of
$\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
unequal spin events. This degeneracy produces straight lines in the parameter
space with negative slopes that depend on the mass ratio, as shown in
Fig.\ref{fig:ps_q148}. Notably, the ansatz performs better in the top left
region than the bottom right along a degeneracy line. In an attempt to address
this issue, we partitioned the parameter space into four regions, as described
in Sec. \ref{sec:result}. However, even with separate optimizations for each
regions, Fig.\ref{fig:ps_q148_quadrant} indicates that the bottom-right region
exhibits similar mismatches as before, while the top-left region's performance
has improved. This observation suggests that the ansatz is specific to certain
regions of the parameter space, with a preference for BBH events featuring
$\chi_1<0$ and $\chi_2>0$.

The division of the parameter space into four regions was a simple approach
taken for practical reasons. A more systematic approach would involve the use of
level set estimation algorithms to identify regions of interest within the
parameter space. Such an algorithm can reveal additional degeneracies or issues
that may exist within the ansatz. One possible strategy is to recalibrate
individual regions of interest to achieve better results. An alternative
approach is to select regions based on degeneracy lines. However, due to the
limited number of NR waveforms available, we were unable to implement this
approach. With more NR waveforms available in the future that cover the entire
parameter space, we can perform optimization with fewer restrictions and select
regions more systematically.

@@

Furthermore, we investigated how the intrinsic parameters affect the
improvement. Fig.~\ref{fig:ps_q148} shows that the optimization procedure has a
certain preference for waveforms lying in the top-left region while the fourth
quadrant cannot be improved. To further test this result, we recalibrate
waveforms in separate regions in parameter space. As shown in
Fig.~\ref{fig:ps_q148_quadrant}, this recalibration process gives further
improvement to the top-left region while the bottom-right region shows similar
result. This indicates that the model's ansatz does not fit waveforms in this region. This phenomenon is due to the reduced spin approximation used
in parameterizing the ansatz, where degeneracies between $\chi_1$ and $\chi_2$
are introduced.

While we naively separate the optimization process into 4 regions, one can
perform systematic region-selection. In principle, we can apply this general
method to other newer and more accurate models such as IMRPhenomX or IMRPhenomP
models. Then, we can perform all the above analyses to understand how to
construct better GW Phenom models in the future.
